### 2020-05-26 writeup/intro.tex
wip

Identifying when negative transfer occurs requires comparing the performance of multi-task learning directly to single-task learning.
The technical challenge is to develop generalization bounds that are able to scale tightly with the qualities of task data, such as the number of datapoints.
In classical Rademacher or VC based theory of multi-task learning \cite{B00,AZ05,M06}, the generalization bounds are usually presented in a way so that the error goes down as more labeled data is added.
On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.
More recent work has shown the benefit of learning multi-task representations for certain half-spaces \cite{MPR16} and multipl sparse regressions \cite{LPTV09,LPVT11}.

@@

%We formulate the heterogeneity between different task data under covariate and model shifts .
%\begin{algorithm}[!t]
%	\caption{Multi-task learning using a hard-parameter sharing architecture}
%	\label{alg_estimator}
%	\begin{algorithmic}[1]
%		\Input Two regression tasks $(X_1, Y_2)$, $(X_2, Y_2)$.
%		\Param Shared body $B$, task-specific prediction heads $W_1, W_2$.
%		\State Training the shared body $B$.
%		\State Optimizing the task heads on the validation set

%		\begin{itemize}
%			\item Jointly optimizing both tasks: $\hat{\beta}_t^{\MTL}$
%			\item Optimizing the target task: $\hat{\beta}_t^{\TL}$
%			\item Single-task training baseline: $\hat{\beta}_t^{\STL}$
%		\end{itemize}
%		\State Problem statement: how can we compare the test error of the three estimators on the target task?
		%, $\te(\hat{\beta}_t^{\MTL})$, $\te(\hat{\beta}_t^{\STL})$ and $\te(\hat{\beta}_t^{\TL})$?
%	\end{algorithmic}
%\end{algorithm}

\textbf{Main results.}
	We begin by observing that $\te(\hat{\beta}_t^{\MTL})$ can be decomposed into two parts, a variance part that is reduced from $\te(\hat{\beta}_t^{\STL})$, and a bias part that captures the difference between $\beta_t$ and the rest of $\beta$'s.
	We term the bias part as \textit{model shift bias}.
	Intuitively, whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ depends on the trade-off between the amount of variance reduced and the model shift bias part.
	For the high-dimensional regime when $p$ goes to infinity, we derive the asymptotic limit of $\te(\hat{\beta}_t^{\MTL}) - \te(\hat{\beta}_t^{\STL})$ as a function of the number of datapoints $\set{n_1, n_2}$, the covariance matrices $\set{\Sigma_1, \Sigma_2}$, the ground truth parameters $\set{\beta_1, \beta_2}$, and a certain fixed value derived from solving equation \eqref{eq_mtl} (see Theorem \ref{thm_main_informal} for the statement).
	Then, we show a similar result for any number of tasks that have the same covariates, i.e. the $X_i$'s are equal to each other.
	This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
	Based on the technical result, we provide theoretical insights over when multi-task learning performs better than single-task learning.

Our first theoretical insight is to explain a sharp transition from positive to negative transfer as a parameter of \textit{task model dissimilarity} $\norm{\beta_1 - \beta_2}^2$.
	In Proposition \ref{prop_dist_transition}, we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.

Our second theoretical insight is to show that performing multi-task learning reduces the need for labeled data, which is a key finding of Taskonomy \cite{ZSSGM18}.
	We define the \textit{data efficiency ratio} as the smallest $\alpha\in(0, 1)$ such that if we only use an $\alpha$ fraction of labeled data, then the test error of $\hat{\beta}_t^{\MTL}$ matches that of the $\hat{\beta}_t^{\STL}$ on the entire set.
	In Proposition \ref{prop_data_efficiency}, we show that the data efficiency ratio of an illustrative example is at most $\alert{\frac {1}{2\rho_t}}$.

Our third theoretical insight is that as the data size becomes more imbalanced between the source and target task, having mis-aligned task covariance matrices can result in suboptimal performance.
	In Proposition \ref{prop_covariate}, we show that as $n_1 / n_2$ becomes large, the best performing source task has have the same covariance matrix as the target task.
	On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.

Finally, while our focus is multi-task learning, our setup and technical tools are also applicable to transfer learning settings.
In Section \ref{sec_taskonomy}, we provide a case study of the transfer function used in Taskonomy \cite{ZSSGM18}.
We observe a similar trade-off between model-shift bias and variance for this setting.

@@

\textbf{Experimental results.} We show practical implications of our theoretical insights on text and image classification tasks.
	First, we provide a metric to determine positive versus negative transfer by comparing the test accuracies of single-task models.
	%validate on text and image classification tasks that comparing single-task accuracies can help determine whether multi-task learning performs better than single-task learning.
	Second, we validate that performing multi-task learning can reduce the need for labeled data.
	On 6 sentiment analysis tasks, we observe that just by using \alert{40\%} of the labeled data, the overall accuracy of multi-task learning matches that of learning every task in isolation.
	Finally, we show that as the data size between the source and target task becomes more imbalanced, aligning the covariances of the tasks becomes more beneficial.
%	On the other hand, if the number of source task datapoints is comparable to the target task, aligning task covariances may hurt performance.

>>>>

Identifying when negative transfer occurs requires comparing the performance of multi-task learning directly to single-task learning.
The technical challenge is to develop generalization bounds that are able to scale tightly with the qualities of task data, such as the number of datapoints.
In classical Rademacher or VC based theory of multi-task learning \cite{B00,AZ05,M06}, the generalization bounds are usually presented in a way so that the error goes down as more labeled data is added.
On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.
More recent work has shown the benefit of learning multi-task representations for certain half-spaces \cite{MPR16} and multiple sparse regressions \cite{LPTV09,LPVT11}.

@@

\textbf{Main results.}
	We begin by observing that $\te(\hat{\beta}_t^{\MTL})$ can be decomposed into two parts, a variance part that is reduced from $\te(\hat{\beta}_t^{\STL})$, and a bias part that captures the difference between $\beta_t$ and the rest of $\beta$'s.
	We term the bias part as \textit{model shift bias}.
	Intuitively, whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ depends on the trade-off between the amount of variance reduced and the model shift bias part.
	For the high-dimensional regime when $p$ goes to infinity, we derive the asymptotic limit of $\te(\hat{\beta}_t^{\MTL}) - \te(\hat{\beta}_t^{\STL})$ as a function of the number of datapoints $\set{n_1, n_2}$, the covariance matrices $\set{\Sigma_1, \Sigma_2}$, the ground truth parameters $\set{\beta_1, \beta_2}$, and a certain fixed value derived from solving equation \eqref{eq_mtl} (see Theorem \ref{thm_main_informal} for the statement).
	Then, we show a similar result for any number of tasks that have the same covariates, i.e. the $X_i$'s are equal to each other.
	This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.

	While our focus is multi-task learning, our setup and technical tools are also applicable to transfer learning settings.
	In Section \ref{sec_taskonomy}, we provide a case study of the transfer function used in Taskonomy \cite{ZSSGM18}.
	We observe a similar trade-off between model-shift bias and variance for this setting.

	Based on the technical result, we provide theoretical insights over when multi-task learning performs better than single-task learning.
	\begin{itemize}
		\item First, we explain a sharp transition from positive to negative transfer as a parameter of \textit{task model dissimilarity} $\norm{\beta_1 - \beta_2}^2$.
	In Proposition \ref{prop_dist_transition}, we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.

		First, we provide a metric to determine positive versus negative transfer by comparing the test accuracies of single-task models.
		\item Second, we show that performing multi-task learning reduces the need for labeled data, which is a key finding of Taskonomy \cite{ZSSGM18}.
	We define the \textit{data efficiency ratio} as the smallest $\alpha\in(0, 1)$ such that if we only use an $\alpha$ fraction of labeled data, then the test error of $\hat{\beta}_t^{\MTL}$ matches that of the $\hat{\beta}_t^{\STL}$ on the entire set.
	In Proposition \ref{prop_data_efficiency}, we show that the data efficiency ratio of an illustrative example is at most $\alert{\frac {1}{2\rho_t}}$.

			Second, we validate that performing multi-task learning can reduce the need for labeled data.
	On 6 sentiment analysis tasks, we observe that just by using \alert{40\%} of the labeled data, the overall accuracy of multi-task learning matches that of learning every task in isolation.
		\item Finally, as the data size becomes more imbalanced between the source and target task, having mis-aligned task covariance matrices can result in suboptimal performance.
		In Proposition \ref{prop_covariate}, we show that as $n_1 / n_2$ becomes large, the best performing source task has have the same covariance matrix as the target task.
		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.

		Finally, we show that as the data size between the source and target task becomes more imbalanced, aligning the covariances of the tasks becomes more beneficial.
	\end{itemize}

@@

%\textbf{Experimental results.} We show practical implications of our theoretical insights on text and image classification tasks.

	%validate on text and image classification tasks that comparing single-task accuracies can help determine whether multi-task learning performs better than single-task learning.

%	On the other hand, if the number of source task datapoints is comparable to the target task, aligning task covariances may hurt performance.

### 2020-05-26 writeup/setup.tex

Recall that we have $t$ labeled tasks available, denoted by $(X_1, Y_1), (X_2, Y_2), \dots, (X_t, Y_t)$.
Following \cite{HMRT19,BLLT20}, we assume that for each task $i = 1,2,\dots,t$,  every feature vector is generated as $x = \Sigma_i^{1/2} z$, where $z\in\real^p$ is a random vector with i.i.d. entries of mean zero and unit variance and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
Without loss of generality, let the $t$-th task denote the target task.
For an estimator $\hat{\beta}\in\real^p$, the test error for target task is defined as
	\begin{align*}
		\te_t(\hat{\beta}) \define \exarg{z}{\exarg{\varepsilon_t}{({(\Sigma_t^{1/2} z)}^{\top}\hat{\beta} - {(\Sigma_t^{1/2})}^{\top}\beta_t)^2}}
		= \exarg{\varepsilon_t}{(\hat{\beta} - \beta_t)^{\top}\Sigma_t(\hat{\beta} - \beta_t)}.
	\end{align*}

Let $\hat{\beta}_t^{\STL} = (X_t^{\top}X_t)^{-1}X_t^{\top}Y_t$ denote the linear regression estimator, which can be also be obtained from equation \eqref{eq_mtl} by using the target task alone.
The bias-variance trade-off \cite{HTF09} says that
	\[ \te_t(\hat{\beta}) =
		\bignorm{\exarg{\varepsilon_t}{\hat{\beta}} - \beta_t}^2 + \exarg{\varepsilon_t}{\hat{\beta} - \exarg{\varepsilon_t}{\hat{\beta}}}. \]
Recall that we assume $n_i > p$ for every $1\le i\le t$.
Hence the bias of $\te_t(\hat{\beta}_t^{\STL})$ is zero and its test error is equal to variance, given by $\sigma^2 \cdot \tr[(X_i^{\top}X_i)^{-1}\Sigma_i]$.
We focus on a setting where the data size $n_t$ of the target task is equal to a small constant $\rho_t > 1$ times $p$.
This setting captures the need for adding more labeled data to reduce the test error of the target task.

%, i.e. %(following the argument of Proposition \ref{prop_monotone}), i.e.
%\[ \bigtr{(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2} \le \bigtr{(X_2^{\top}X_2)^{-1}\Sigma_2}. \]
%Because of model shift bias, we can no longer guarantee that $\te(\hat{\beta}_{t}^{\MTL}) \le \te(\hat{\beta}_t^{\STL})$.

>>>>

Recall that we have $t$ labeled tasks available, denoted by $(X_1, Y_1), (X_2, Y_2), \dots, (X_t, Y_t)$.
Following \cite{HMRT19,BLLT20}, we assume that for each task $i = 1,2,\dots,t$,  every feature vector is generated as $x = \Sigma_i^{1/2} z$, where $z\in\real^p$ is a random vector with i.i.d. entries of mean zero and unit variance and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
Without loss of generality, let the $t$-th task denote the target task.
For an estimator $\hat{\beta}\in\real^p$, the test error of the target task is defined as
	\begin{align*}
		\te_t(\hat{\beta}) \define \exarg{z}{\exarg{\varepsilon_t}{({(\Sigma_t^{1/2} z)}^{\top}\hat{\beta} - {(\Sigma_t^{1/2})}^{\top}\beta_t)^2}}
		= \exarg{\varepsilon_t}{(\hat{\beta} - \beta_t)^{\top}\Sigma_t(\hat{\beta} - \beta_t)}.
	\end{align*}

Let $\hat{\beta}_t^{\STL} = (X_t^{\top}X_t)^{-1}X_t^{\top}Y_t$ denote the linear regression estimator, which can be also be obtained from equation \eqref{eq_mtl} by using the target task alone.
The bias-variance trade-off \cite{HTF09} says that
	\[ \te_t(\hat{\beta}) =
		\bignorm{\exarg{\varepsilon_t}{\hat{\beta}} - \beta_t}^2 + \exarg{\varepsilon_t}{\hat{\beta} - \exarg{\varepsilon_t}{\hat{\beta}}}. \]
Recall that $n_i = \rho_i \cdot p$ where $\rho_i > 1$ is a fixed constant for every $1\le i\le t$.
Hence the bias of $\te_t(\hat{\beta}_t^{\STL})$ is zero and its test error is equal to variance, given by $\sigma^2 \cdot \tr[(X_i^{\top}X_i)^{-1}\Sigma_i]$.
We focus on a setting where the data size $n_t$ of the target task is equal to a small constant $\rho_t > 1$ times $p$.
This setting captures the need for adding more labeled data to reduce the test error of the target task.

In order to study the trade-off between model-shift bias and variance reduction, we need tight concentration bounds to quantify both effects.
For this purpose, we consider the high-dimensional setting where $p$ goes to infinity.
A well-known result for this setting states that the asymptotic limit of $\te(\hat{\beta}_t^{\STL})$ is $\frac {\sigma^2} {\rho_t - 1}$ (e.g. Chapter 6 of \cite{S07}), which scales with the data size and noise level of the target task.
However, this result only applies to the single-task setting.
Therefore, we focus on deriving the asymptotic limit of $\te(\hat{\beta}_t^{\MTL})$ when $p$ goes to infinity.
%, i.e. %(following the argument of Proposition \ref{prop_monotone}), i.e.
%\[ \bigtr{(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2} \le \bigtr{(X_2^{\top}X_2)^{-1}\Sigma_2}. \]
%Because of model shift bias, we can no longer guarantee that $\te(\hat{\beta}_{t}^{\MTL}) \le \te(\hat{\beta}_t^{\STL})$.

To illustrate our intuition, we begin by considering the multi-task learning setting with one source task.
Recall that $\hat{\beta}_t^{\MTL}$ is defined as $BW_t$ after solving equation \eqref{eq_mtl}.
We decompose the test error of $\hat{\beta}_{t}^{\MTL}$ on the targe task into two parts (to be derived in Appendix \ref{app_proof_main})
\begin{align}
	\te_t(\hat{\beta}_t^{\MTL}) =& ~ \hat{v}^2 \bignorm{\Sigma_2^{1/2} (\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}X_1^{\top}X_1 (\beta_1 - \hat{v}\beta_2)}^2 \label{eq_te_model_shift} \\
	&+ \sigma^2\cdot \bigtr{(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2}, \label{eq_te_var}
\end{align}
where $\hat{v} = W_2 / W_1$ denotes the ratio of the output layer weights.
